---
title: "Gapminder Data - Exploratory Analysis"
author: "Luc Frachon"
date: "12 janvier 2017"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This study is a Problem Set of the "Data Analysis with R" course, part of Udacity's Data Analyst Nanodegree curriculum. In this study, we select datasets from [Gapminder](www.gapminder.org/data) and perform exploratory analyses using different types of visualisations.  

For educational purposes, we make the code visible. The code is developped using R and the following libraries: tidyr, dplyr, ggplot2, gridExtra and RColorBrewer.

-----

# 1. Life Expectancy at Birth over Time

## Questions

The questions we would like to answer are the following:
 
 * How did Life Expectancy at Birth (LEB) and its distribution among countries evolve overall since the middle of the 20th century?
 * What are the periods where it has changed the most?
 * How did the evolution of LEB compare between countries of different wealth?

## Data Load

We select the Life Expectancy At Birth dataset. We download the file, save it as a .tsv ("tabulation-separated values") and load it into R:

```{r cache = FALSE}
require(tidyr)
require(dplyr)
require(ggplot2)
require(gridExtra)
require(RColorBrewer)

# Load the data file about Life Expectancy:
le_full <- read.csv('./indicator_life_expectancy_at_birth.tsv', sep = '\t', 
                stringsAsFactors = FALSE)

```

Now let's take a first look at the data:
```{r}
colnames(le_full)[1] <- "country"
le_full[1:5, 1:5]

```


The data documentation specifies that data prior to 1950 is estimated from models using other data, and might thus be unreliable. We will therefore only use post-1950 data.
Let us look at the data structure after this:

```{r}
le <- select(le_full, country, num_range("X", 1950:2016))
dim(le)
le[1:5, 1:5]
```

The data contains 260 countries and covers 67 years (+ the "country" column).

## LEB distribution in 1950 vs. 2016

Let's look at some summary statistics and plot the data distribution for 1950 and 2016 (first and latest year in the sample):
```{r}
# Some summary statistics for 1950 and 2016:
print("The World in 1950:")
summary(le$X1950)
print(paste("Standard deviation:", sd(le$X1950, na.rm = TRUE)))

print("The World in 2016:")
summary(le$X2016)
print(paste("Standard deviation:", sd(le$X2016, na.rm = TRUE)))

# Histograms for both years:
p1 <- ggplot(aes(x = X1950), data = le) +
    theme_dark() +
    geom_histogram(binwidth = 1, color = 'black', fill = 'steelblue') +
    ggtitle("Life Expectancy Distribution in 1950") + 
    scale_x_continuous(limits = c(0, 100))
p2 <- ggplot(aes(x = X2016), data = le) +
    theme_dark() +
    geom_histogram(binwidth = 1, color = 'black', fill = 'steelblue') +
    ggtitle("Life Expectancy Distribution in 2016") +
    scale_x_continuous(limits = c(0, 100))
grid.arrange(p1, p2, ncol = 1)

ggsave('hist_1950_2016.png')


```

These numbers and plots show that the modern world is a radically different place to 1950. The mean LBE moved from 49.3 to 72.6 years while the minimum LBE more than doubled from 23.5 to 48.9. At the same time, the standard deviation $\sigma$ decreased from 11.8 to 7.7, meaning that not only people live longer on average, but there is also less dissimilarities between countries.

These trends are obvious from the histograms. The whole plot shifted to the right between 1950 and 2016. The data from 1950 is more spread out and is somewhat bi-modal, hinting at well-separated groups of countries. It exhibits almost no skewness.  
In contrast, the 2016 data is unimodal, less spread out and negatively skewed (the left tail is longer than the right tail). In other words, the majority of countries are above the average -- an observation confirmed by the fact that median > mean.

Let us see if any countries have seen their LEB decrease:

```{r}
le_decrease <- filter(le, X2016 < X1950)
nrow(le_decrease)
```

Every single country has had improvements to their LEB.

# Evolution of LEB over time

To get a better sense for the rate of improvement in LEB, let us select a few representative milestones and compare their LEB distributions accross countries. We will focus on years 1950, 1970, 1990, 2010 and 2016.

## Frequency Polygons

We plot the distribution for each of the selected years using frequency polygons, which allow us to represent several data vectors on the same plot.


```{r}
# Convert data to long format:
le_long <- gather(le, key = year, value = life_exp, -country, convert = TRUE)
le_long$year <- as.factor(sub('X', '', le_long$year))

# Subset 15-year periods:
fifteen_y <- as.character(seq(1956, 2016, 15))
le_fifteen <- filter(le_long, year %in% fifteen_y)
```

```{r}
ggplot(aes(x = life_exp), data = le_fifteen) +
    theme_dark() +
    geom_freqpoly(aes(color = year), binwidth = 5, size = 1) +
    scale_color_brewer(type = 'seq', palette = 'Blues')
ggsave('freqpoly_1950-2016_by_decade.png')
```

It would appear that the biggest changes happened in the earlier years of the dataset (up to 1986), with the distribution's shape shifting from bimodal to unimodal and negative skeweness. The rate of change seems comparatively slower in the years after that

Let's look at some boxplots for a more quantitative view:


```{r}
five_y <- as.character(seq(1951, 2016, 5))
le_five <- filter(le_long, year %in% five_y)
ggplot(aes(x = year, y = life_exp), data = le_five) +
    theme_dark() +
    geom_boxplot(fill = "lightsteelblue") + 
    ggtitle("Life Expectancy Worldwide, 1950-2016")
ggsave('box_1950-2016_by_decade.png')
```

 * As noted before, the largest improvements in median values happened between approximately 1950 and 1980,
 * The growth seemed to slow down the the late 80's / early 90's, then bounced back especially in the early 2000's,
 * The interquartile range (IQR -- spread of the middle 50% of the countries) decreased most significantly from 1971 to 1986,
 * The median progressively shifted towards the upper bound of the IQR, confirming a shift towards negative skewness
 * A few outliers appeared aound 2006 at the lower end of the distribution.
 
 
## Evolution by Country Wealth

The shift from a bimodal distribution to unimodal with negative skewness might be hint to the fact that there were well-defined groups of countries (in terms of life expectancy) that progressively merged while improving. Typically, poor countries will tend to have lower life expectancy ([see here for an illustration of this](http://www.gapminder.org/world/#$majorMode=chart$is;shi=t;ly=2003;lb=f;il=t;fs=11;al=30;stl=t;st=t;nsl=t;se=t$wst;tts=C$ts;sp=5.59290322580644;ti=2013$zpv;v=0$inc_x;mmid=XCOORDS;iid=phAwcNAVuyj1jiMAkmq1iMg;by=ind$inc_y;mmid=YCOORDS;iid=phAwcNAVuyj2tPLxKvvnNPA;by=ind$inc_s;uniValue=8.21;iid=phAwcNAVuyj0XOoBL_n5tAQ;by=ind$inc_c;uniValue=255;gid=CATID0;by=grp$map_x;scale=log;dataMin=194;dataMax=96846$map_y;scale=lin;dataMin=23;dataMax=86$map_s;sma=49;smi=2.65$cd;bd=0$inds=;modified=60)), so we are interested in comparing the rate of improvement for different groups of countries defined by their GDP per capita. The suspicion is that poor countries improved their LEB faster than rich ones, and that this catching up explains the previous observations.

For a full analysis of potential correlations, we would need to join the LEB data with GDP data but in this exploratory analysis, we just want to get a feel for the correctness of our assumption. We will therefore select only a few countries in each group.

*Note:* The countries are classified according to their GDP per capita as of 2015.

With the defined groups, we can then build boxplots for each, similar to the work done above.


```{r fig.height=8, fig.width=7}

# Define country groups:
poor <- c("Somalia", "Congo, Dem. Rep.", "Mozambique", "Ethiopia")
mid_lower <- c("Bangladesh", "Kenya", "Cameroon", "India")
mid_upper <- c("Indonesia", "China", "Brazil", "Russia")
rich <- c("United States", "Japan", "Switzerland", "United Kingdom")

le_5_select <- subset(le_five, 
                        country %in% c(poor, mid_lower, mid_upper, rich))
le_5_select$country_group <- NA
le_5_select[le_5_select$country %in% poor, 
              "country_group"] <- "poor"
le_5_select[le_5_select$country %in% mid_lower, 
              "country_group"] <- "mid_lower"
le_5_select[le_5_select$country %in% mid_upper, 
              "country_group"] <- "mid_upper"
le_5_select[le_5_select$country %in% rich, 
              "country_group"] <- "rich"

# Build boxplots:
p1 <- ggplot(aes(x = as.factor(year), y = life_exp), 
             data = subset(le_5_select, country_group == "poor")) +
      theme_dark() +
      geom_boxplot(fill = 'lightsteelblue') +
      coord_cartesian(ylim = c(30, 85)) +
      ggtitle("Low GDP per Capita") +
      xlab("Year") + ylab("Life Expectancy at Birth") +
      theme(axis.text.x = element_text(angle = 90, hjust = 1))

p2 <- ggplot(aes(x = as.factor(year), y = life_exp), 
      data = subset(le_5_select, country_group == "mid_lower")) +
      theme_dark() +
      geom_boxplot(fill = 'lightsteelblue') +
      coord_cartesian(ylim = c(30, 85)) +
      ggtitle("Lower-Mid GDP per Capita") +
      xlab("Year") + ylab("Life Expectancy at Birth") +
      theme(axis.text.x = element_text(angle = 90, hjust = 1))
      
p3 <- ggplot(aes(x = as.factor(year), y = life_exp), 
      data = subset(le_5_select, country_group == "mid_upper")) +
      theme_dark() +
      geom_boxplot(fill = 'lightsteelblue') +
      coord_cartesian(ylim = c(30, 85)) +
      ggtitle("Upper-Mid GDP per Capita") +
      xlab("Year") + ylab("Life Expectancy at Birth") +
      theme(axis.text.x = element_text(angle = 90, hjust = 1))

p4 <- ggplot(aes(x = as.factor(year), y = life_exp), 
      data = subset(le_5_select, country_group == "rich")) +
      theme_dark() +
      geom_boxplot(fill = 'lightsteelblue') +
      coord_cartesian(ylim = c(30, 85)) +
      ggtitle("High GDP per Capita") +
      xlab("Year") + ylab("Life Expectancy at Birth") + 
      theme(axis.text.x = element_text(angle = 90, hjust = 1))

grid.arrange(p1, p2, p3, p4, ncol = 2)
ggsave('box_1950-2016_by_decade_by_group.png')
```

These plots tend to confirm our assumption. IQRs are not very relevant with such small samples, but we can clearly see that medians improved significantly more for countries in the "Low GDP", "Lower-Mid GDP" and "Upper-Mid" groups (by about 25 to 27 years for each of them), than the "High GDP" group (about 14 years). Moreover, it seems that the richer the country, the slower the rate of change between 2010 and 2016.

These trends make sense, as the law of diminishing returns would apply to life expectancy -- it takes a lot more effort to get from 60 to 80 than to get from 40 to 60. However, they also indicate that health in developping countries have improved dramatically in the second half of the $XX^{th}$ century and they continue to improve.

## Recap: Observations on Life Expectancy at Birth

This exploratory analysis has given us some key insights into the data. We observed that the overall LEB has improved dramatically since 1950 for every single country in the world. This improvement has been faster between 1950 and 1990 than in subsequent years, but it is still increasing. More importantly, the improvement benefited poorer countries the most, which led to a much more compact distribution of LEB accross the world.

A few additional ideas for analyses arise from this preliminary work:

 * Conduct a more complete detailed of the rate of change vs. GDP per capita,
 * Group coutries by continent,
 * Look for outliers in the rate of change and see if historical events can explain them.

-----

# 2. Life Expectancy vs DTP3 Immunization

In this second part of the analysis, we introduce a second variable, the proportion of one-year-olds immunized with 3 doses of Diphteria-Tetanus-Pertussis (DTP3). This dataset contains observations since 1980 and is also provided by [Gapminder](www.gapminder.org/data).

## Questions:

We would like to answer the following questions:

 * How did the immunization rate evolve overall and in specific countries in the last 36 years?
 * Is there a correlation between the immunization rate and life expectancy at birth?
 
## Data Load

```{r}
dtp <- read.csv('./one_year_old_dtp3_immunized.tsv', sep = '\t',
                stringsAsFactors = FALSE)
colnames(dtp)[1] <- 'country'
dim(dtp)
```

The data contains 270 countries over 32 years (from 1980 to 2011). Note that the LEB dataset previously explored had only 260 countries but spanned a much longer period. To compare the two, we will need to abandon some of the data.

## Evolution of DTP3 Immunization over Time

We first need to convert the dataset into a long format:

```{r}
dtp_long <- gather(dtp, key = year, value = pct_immunized, 
                   -country, convert = TRUE)
dtp_long$year <- as.factor(sub('X', '', dtp_long$year))
```

Let's build the same frequency polygon plot as with the previous dataset, selecting a few years:

```{r}
ggplot(aes(x = pct_immunized), data = subset(dtp_long, year %in% c('1981', '1991', '2001', '2011'))) +
    theme_dark() +
    geom_freqpoly(aes(color = year), binwidth = 5, size = 1) +
    scale_color_brewer(type = 'seq', palette = 'Blues') +
    scale_x_continuous(limits = c(0, 100))
```

While in 1981, the distribution of countries was very uniform, by 1991 it was already starting to display a significant peak between 80 and 100%. This trend continued in the next 20 years until the distribution looked heavily negatively skewed, with a left-tail comprising of a only a few countries.

```{r}

decades <- as.character(seq(1981, 2011, 2))
ggplot(aes(x = year, y = pct_immunized), 
       data = subset(dtp_long, year %in% decades)) +
    theme_dark() +
    geom_boxplot(fill = "lightsteelblue") + 
    ggtitle("Percentage of 1-year-old immunized with DTP3")
```

Indeed, these boxplots confirm the massive increase in immunization and the equally critical reduction of the distribution's dispersion.

Finally, let's re-use the country groups created in Part 1 to compare how DTP3 Immunization progressed between countries of different wealth.

```{r fig.height=8, fig.width=7}
dtp_select <- subset(dtp_long, 
                        country %in% c(poor, mid_lower, mid_upper, rich))
dtp_select$country_group <- NA
dtp_select[dtp_select$country %in% poor, 
              "country_group"] <- "poor"
dtp_select[dtp_select$country %in% mid_lower, 
              "country_group"] <- "mid_lower"
dtp_select[dtp_select$country %in% mid_upper, 
              "country_group"] <- "mid_upper"
dtp_select[dtp_select$country %in% rich, 
              "country_group"] <- "rich"

# Build boxplots:
p1 <- ggplot(aes(x = year, y = pct_immunized), 
             data = subset(dtp_select, 
                           year %in% decades & country_group == "poor")) +
      theme_dark() +
      geom_boxplot(fill = 'lightsteelblue') +
      coord_cartesian(ylim = c(0, 100)) +
      ggtitle("Low GDP per Capita") +
      xlab("Year") + ylab("% immunized with DTP3")

p2 <- ggplot(aes(x = year, y = pct_immunized), 
      data = subset(dtp_select, 
                    year %in% decades & country_group == "mid_lower")) +
      theme_dark() +
      geom_boxplot(fill = 'lightsteelblue') +
      coord_cartesian(ylim = c(0, 100)) +
      ggtitle("Lower-Mid GDP per Capita") +
      xlab("Year") + ylab("% immunized with DTP3")
      
p3 <- ggplot(aes(x = year, y = pct_immunized), 
      data = subset(dtp_select, 
                    year %in% decades & country_group == "mid_upper")) +
      theme_dark() +
      geom_boxplot(fill = 'lightsteelblue') +
      coord_cartesian(ylim = c(0, 100)) +
      ggtitle("Upper-Mid GDP per Capita") +
      xlab("Year") + ylab("% immunized with DTP3")

p4 <- ggplot(aes(x = year, y = pct_immunized), 
      data = subset(dtp_select, 
                    year %in% decades & country_group == "rich")) +
      theme_dark() +
      geom_boxplot(fill = 'lightsteelblue') +
      coord_cartesian(ylim = c(0, 100)) +
      ggtitle("High GDP per Capita") +
      xlab("Year") + ylab("% immunized with DTP3")

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

As with Life Expectancy at Birth, the biggest gains happened in the poorer countries. The lowest group however is still far from the levels reached by the other countries and still has more dispersion.

## Life Expectancy at Birth vs. DTP3 Immunization of 1-year-olds

In this section, we will explore possible correlations between Life Expectancy at Birth and DTP3 immunization among 1-year-old infants.

We first have to join the two datasets. As we already noticed that the LEB dataset had less countries but more years that the DTP3 dataset, we will need to subset it for the same time span and then perform a left join.
```{r}
period <- as.character(seq(1980, 2011))
le_dtp <- le %>% 
    filter(year %in% period) %>%
    left_join(dtp_long, by = c("country", "year"))


```


